Evaluating `Graphical Perception' with CNNs

### 2018-08-15 PAPER/04_position_angle_experiment.tex
Update paper from overleaf.

\clearpage
\section{Experiment: Position-Angle}

Cleveland and McGill measure human perception to the ratios of positions and angles through comparisons on bar charts and pie charts~\cite{cleveland_mcgill}. We create rasterized images following Cleveland and McGill's proposed encoding and investigate computational perception of our networks (Figure~\ref{fig:teaser}). These have five bar or pie sectors representing numbers which add to 100, where each is greater than three and smaller than 39. One required change is in the minimal differences between the values: Cleveland and McGill create stimuli where the differences between each number are greater than $0.1$. However, as our networks only take $100\times100$ pixel images as input, we can only minimally represent a difference of $1$ pixel.

@@

%\begin{figure}[t]
%	  \includegraphics[width=\linewidth]{figure3_overview}
%  \caption{\textbf{Position-Angle Experiment.} We create rasterized visualizations of pie charts and bar charts to follow Cleveland and McGill's position-angle experiment. The experimental task involves the judgement of different encoded values in comparison to the largest encoded values. The pie chart (left) and the bar chart (right) visualize the same data point. In their paper, Cleveland and McGill report less errors using bar charts.}
%	\label{fig:position_angle_experiment}
%\end{figure}
%\begin{table}[h]
%\centering
%\caption{\textbf{Position-Angle Experiment.} We create rasterized visualizations of pie charts and bar charts to follow Cleveland and McGill's position-angle experiment. The experimental task involves the judgement of different encoded values in comparison to the largest encoded values. The pie chart and the bar chart visualize the same data point. In their paper, Cleveland and McGill report less errors using bar charts.}
%\resizebox{\linewidth}{!}{
%\begin{tabular}{lllr}
%	\toprule
%	\multicolumn{2}{l}{~} & ~ & Permutations\\
%
%	\midrule
%	\raisebox{-.85\height}{\includegraphics[width=.5in]{figure3_Bar_Chart.pdf}} & \makecell[tl]{Type 1: \emph{Bar Chart}\\~~~Perceptual Task: \emph{Position}\\~ \\~ \\} &~& \makecell[tr]{~\\ $878,520$}\\
%
%
%	\midrule
%	\raisebox{-.85\height}{\includegraphics[width=.5in]{figure3_Pie_Chart.pdf}} & \makecell[tl]{Type 2: \emph{Pie Chart}\\~~~Perceptual Task: \emph{Angle}\\~ \\~ \\} &~& \makecell[tr]{~\\ $878,520$}\\
%
%
%	\bottomrule
%\end{tabular}
%}
%\label{tab:pos_angle_parameters}
%\end{table}

@@

\begin{hypolist}
	\item \textbf{H2.1} \textbf{Computed perceptual accuracy will be higher for bar charts than pie charts.} Cleveland and McGill report that position judgements are almost twice as accurate (MLAE) as angle judgements in humans. Following our ranking of elementary perceptual tasks (Table~\ref{tab:ranking}), we see that our networks also judge position encodings more accurately than angles, and so our networks will be able to more easily judge bar charts than pie charts.
	\item \textbf{H2.2} \textbf{Convolutional neural networks will learn to regress bar chart ratios faster than pie chart ratios in training.} This follows directly from H2.1.
\end{hypolist}

@@

\begin{figure}[t]
	\centering
	  \includegraphics[width=\linewidth]{figure3_mlae_better_all.pdf}
  \caption{\textbf{Computational results of the position-angle experiment.} \textit{Left:} Our encodings of one data point as a pie chart and a bar chart. \textit{Right:} MLAE and 95\% confidence intervals for different networks. VGG19 * and Xception * are using ImageNet weights while all other networks were trained on the stimuli. We train all networks 12 times (4 times for VGG19 and Xception due to significantly longer training times). VGG19 * and Xception * use ImageNet weights. Our results align with Cleveland and McGill's human results, shown in black~\cite{cleveland_mcgill}.}
	\label{fig:figure3_mlae}
\end{figure}

@@

\noindent{\textbf{Perceptual Accuracy.}} Our networks are able to regress the task ratios for bar charts and pie charts (Fig.~\ref{fig:figure3_mlae}).  Cross-validation yields an average $MLAE=2.176$ ($SD=0.456$) for bar charts, and an average $MLAE=3.296$ ($SD=0.77$) for pie charts. This difference is statistically significant ($F_{1,110}=86.061, p<0.01$), and so we \textbf{accept H2.1}.

Post hoc comparisons show that this holds for most networks: 
MLP for pie charts $4.09$ ($SD=0.027$) and for bar charts $2.494$ ($0.068$) is significant ($t_{22}=72.300,p<0.01$);
LeNet for pie charts $ 3.556 $ ($SD= 0.022 $) and for bar charts $ 1.902 $ ($SD= 0.08 $) is significant $t_{22}=66.111, p<0.01$;
VGG19 * with ImageNet weights for pie charts $ 3.561 $ ($SD= 0.047 $) and for bar charts $ 2.601 $ ($SD= 0.113 $) is significant $t_{22}=25.919,p<0.01$;
Xception * with ImageNet weights for pie charts  $ 3.094 $ ($SD= 0.046 $) and for bar charts $ 2.315 $ ($SD= 0.032 $) is significant $t_{22}=46.329,p<0.01$;
Xception from scratch for pie charts $ 1.939 $ ($SD= 0.1 $) and for bar charts $ 1.375 $ ($SD= 0.062 $) is significant $t_{22}=8.276,p<0.01$; but
the difference for VGG19 from scratch (pie charts $ 1.297 $ ($SD= 0.129 $), bar charts $ 1.153 $ ($SD= 0.09 $)) was not significant with $p<0.05$. 
This outcome is in line with the elementary perceptual task results (Table~\ref{tab:ranking}), where VGG19 was the most successful network, where networks trained from scratch were more performant, and where angle was more difficult to learn than position.
\\~\\
\noindent{\textbf{Training Efficiency.} We measure the MSE loss for all networks on previously-unseen validation data during training. We count a network as converged when this validation loss does not decrease after 10 sequential epochs. Figure~\ref{fig:figure3_val_loss} shows this MSE validation loss during the first twenty epochs for each condition, plotted across all cross-validation splits with overdrawn lines. The pie chart loss decreases more slowly, with the average loss over epochs being $0.052$ ($SD=0.015$) for pie charts and $0.037$ ($SD=0.018$) for bar charts. This difference is statistically significant ($F_{1,2238}=20.656, p<0.01$). Thus, we \textbf{accept H2.2}.
\\~\\
To all our networks, the bar chart is a superior visual encoding than a pie chart, in terms of accuracy and efficiency. Cleveland and McGill observe the same effect for accuracy during their human experiments.

@@

\begin{figure}[t]
	  \includegraphics[width=\linewidth]{figure3_val_loss.pdf}
  \caption{\textbf{Training efficiency of the position-angle experiment.} Mean Squared Error (MSE) loss after each epoch during training, computed on previously-unseen validation data. We train all networks 12 times (4 times for VGG19 and Xception due to significantly longer training times). VGG19 * and Xception * use ImageNet weights. All networks converge faster when learning bar charts.}
	\label{fig:figure3_val_loss}
\end{figure}

>>>>

\section{Experiment 2: Position-Angle}
\label{sec:positionangle}

Cleveland and McGill measure how humans perceive the ratios of positions and angles through comparisons on bar charts and pie charts~\cite{cleveland_mcgill}. We create rasterized images following Cleveland and McGill's proposed encoding and investigate computational perception of our networks (Fig.~\ref{fig:teaser}). These have five bar or pie sectors representing numbers that add to 100, where each is greater than three and smaller than 39. One required change is in the minimal differences between the values: Cleveland and McGill create stimuli with minimum scale difference of $0.1$. However, as our networks only take $100\times100$ pixel images as input, we can only minimally represent a difference of $1$ pixel.

@@

%\begin{figure}[t]
%	  \includegraphics[width=\linewidth]{figure3_overview}
%  \caption{\textbf{Position-Angle Experiment.} We create rasterized visualizations of pie charts and bar charts to follow Cleveland and McGill's position-angle experiment. The experimental task involves the judgment of different encoded values in comparison to the largest encoded values. The pie chart (left) and the bar chart (right) visualize the same data point. In their paper, Cleveland and McGill report less errors using bar charts.}
%	\label{fig:position_angle_experiment}
%\end{figure}
%\begin{table}[h]
%\centering
%\caption{\textbf{Position-Angle Experiment.} We create rasterized visualizations of pie charts and bar charts to follow Cleveland and McGill's position-angle experiment. The experimental task involves the judgment of different encoded values in comparison to the largest encoded values. The pie chart and the bar chart visualize the same data point. In their paper, Cleveland and McGill report less errors using bar charts.}
%\resizebox{\linewidth}{!}{
%\begin{tabular}{lllr}
%	\toprule
%	\multicolumn{2}{l}{~} & ~ & Permutations\\
%
%	\midrule
%	\raisebox{-.85\height}{\includegraphics[width=.5in]{figure3_Bar_Chart.pdf}} & \makecell[tl]{Type 1: \emph{Bar Chart}\\~~~Perceptual Task: \emph{Position}\\~ \\~ \\} &~& \makecell[tr]{~\\ $878,520$}\\
%
%
%	\midrule
%	\raisebox{-.85\height}{\includegraphics[width=.5in]{figure3_Pie_Chart.pdf}} & \makecell[tl]{Type 2: \emph{Pie Chart}\\~~~Perceptual Task: \emph{Angle}\\~ \\~ \\} &~& \makecell[tr]{~\\ $878,520$}\\
%
%
%	\bottomrule
%\end{tabular}
%}
%\label{tab:pos_angle_parameters}
%\end{table}

@@

\begin{hypolist}
	\item \textbf{H2.1} \textbf{Computed perceptual accuracy will be higher for bar charts than pie charts.} Cleveland and McGill report that position judgments are almost twice as accurate (MLAE) as angle judgments in humans. Following our ranking of elementary perceptual tasks (Table~\ref{tab:ranking}), we see that our networks also judge position encodings more accurately than angles, and so our networks will be able to more easily judge bar charts than pie charts.
	\item \textbf{H2.2} \textbf{Convolutional neural networks will learn to regress bar chart ratios faster than pie chart ratios in training.} This follows directly from H2.1.
\end{hypolist}

@@

\begin{figure}[tb]
	\centering
	  \includegraphics[width=0.90\linewidth]{figure3_mlae_better_all.pdf}
  \caption{\textbf{Computational results of the position-angle experiment.} \textit{Left:} \change{Example stimuli.} \textit{Right:} MLAE and \change{bootstrapped} 95\% confidence intervals (the lower, the better). VGG19 * and Xception * fine tune ImageNet weights, with all other networks trained from scratch. We show Cleveland and McGill's human results in black~\cite{cleveland_mcgill}.} % We train all networks 12 times (4 times for VGG19 and Xception due to significantly longer training times). 
	\label{fig:figure3_mlae}
\end{figure}

@@

%\JT{State random performance.}

\noindent{\textbf{Perceptual Performance.}} \change{Midmean random performance is at $MLAE=4.7$.} Our networks are able to regress the task ratios for bar charts and pie charts (Fig.~\ref{fig:figure3_mlae}).  Cross-validation yields an average $MLAE=2.176$ ($SD=0.456$) for bar charts, and an average $MLAE=3.296$ ($SD=0.77$) for pie charts. This difference is statistically significant ($F_{1,110}=86.061, p<0.01$), and so we \textbf{accept H2.1}.

Post hoc comparisons show that this holds for most networks: 
MLP for pie charts $4.09$ ($SD=0.027$) and for bar charts $2.494$ ($0.068$) is significant ($t_{22}=72.300,p<0.01$);
LeNet for pie charts $ 3.556 $ ($SD= 0.022 $) and for bar charts $ 1.902 $ ($SD= 0.08 $) is significant $t_{22}=66.111, p<0.01$;
VGG19 * with ImageNet weights for pie charts $ 3.561 $ ($SD= 0.047 $) and for bar charts $ 2.601 $ ($SD= 0.113 $) is significant $t_{22}=25.919,p<0.01$;
Xception * with ImageNet weights for pie charts  $ 3.094 $ ($SD= 0.046 $) and for bar charts $ 2.315 $ ($SD= 0.032 $) is significant $t_{22}=46.329,p<0.01$;
Xception from scratch for pie charts $ 1.939 $ ($SD= 0.1 $) and for bar charts $ 1.375 $ ($SD= 0.062 $) is significant $t_{22}=8.276,p<0.01$; but
the difference for VGG19 from scratch (pie charts $ 1.297 $ ($SD= 0.129 $), bar charts $ 1.153 $ ($SD= 0.09 $)) was not significant with $p<0.05$. 
This outcome is in line with the elementary perceptual task results (Table~\ref{tab:ranking}), where VGG19 was most successful, where networks trained from scratch were more performant, and where angle was more difficult to learn than position.
\\~\\
\noindent{\textbf{Training Efficiency.}} We measure the MSE loss for all networks on previously-unseen validation data during training. Figure~\ref{fig:figure3_val_loss} shows the first twenty epochs for each condition, plotted across all cross-validation splits with overdrawn lines. The pie chart loss decreases more slowly, with the average loss over epochs being $0.052$ ($SD=0.015$) for pie charts and $0.037$ ($SD=0.018$) for bar charts. This difference is statistically significant ($F_{1,2238}=20.656, p<0.01$). Thus, we \textbf{accept H2.2}.
\\~\\
To all our networks, the bar chart is a superior encoding than a pie chart, in terms of accuracy and efficiency. Cleveland and McGill observe the same effect for accuracy during their human experiments.

@@

\begin{figure}[tb]
	  \includegraphics[width=\linewidth]{figure3_val_loss.pdf}
  \caption{\textbf{Training efficiency of the position-angle experiment.} Mean Squared Error (MSE) loss after each epoch during training, computed on previously-unseen validation data. We train all networks 12 times (4 times for VGG19 and Xception due to significantly longer training times). VGG19 * and Xception * use ImageNet weights. All networks reduce MSE loss faster when learning bar charts compared to  pie charts.}
	\label{fig:figure3_val_loss}
\end{figure}

### 2018-08-15 PAPER/05_position_length_experiment.tex
Update paper from overleaf.

\section{Experiment: Position-Length}

The position-length experiment by Cleveland and McGill involves the perception of five types of visualizations~\cite{cleveland_mcgill}. The visualizations are either grouped bar charts or divided bar charts (Fig.~\ref{fig:figure4_mlae}. Both types of graphs can show the same information but the elementary perceptual task is different. According to the theory by Cleveland and McGill, a grouped bar chart always involves the judgement of positions along a common scale. On the other hand, a divided bar chart might require length judgements in addition. Figure~\ref{fig:figure4_mlae} shows different charts of both types on the left. As identified by Cleveland and McGill, types 1, 2, and 3 involve the judgement of positions along a common scale while types 4 and 5 involve the measure of length. This means that this experiment does not just compare grouped versus divided bar charts but rather comparing position and length judgements.

For data generation we follow the same approach as in the original experiment. We generate pairs from ten values generated using Cleveland and McGill's equation

\begin{equation}
s_i = 10\times10^{(i-1)/12},\textnormal{~~~~~}i=1,...,10,
\end{equation}

with two different values for each pair. We then generate 8 other random values in the range of 10 and 93. These boundaries were chosen because we create rasterized visualizations using $100\times100$ pixel and want to stay in frame for a convolutional filter size of $5\times5$ in LeNet. We then visually encode the ten values as type 1--5. The paired quantities get marked by a single pixel. These are our quantities of interest and the task is to estimate the ratio of the smaller to the larger. We model this task as a single value regression. For type 4, we follow Cleveland and McGill's constraint that neither the top or the bottom of the marked quantities match to force length estimations rather than position.

We model this experiment as a single value regression task and ask networks to estimate the percentage of the smaller to the larger marked quantity in each visual encoding. The network first has to find the two marked quantities, then identify the smaller one, and finally estimate the ratio in comparison to the larger quantitiy. The 8 random quantities (which should be ignored by the network) push the number of possible permutations to a massive $9.20E+16$ and creates a very challenging problem.

Finally, we also train `multi' networks which include all five types resulting in an even bigger challenge.

@@

We proposed two hypotheses entering the elementary perceptual task experiment:

\begin{itemize}
  \item \textbf{H3.1} \textbf{Our networks can estimate all types equally well.} A grouped bar chart involves judging a position while a divided bar chart most likely (if not type 2) requires length judgements. Our rankings of elementary perceptual tasks do not yield a strong preference for either across all networks.
  \item \textbf{H3.2} \textbf{Trained `multi' networks work as well as individual trained networks.} Convolutional neural networks have massive numbers of trainable parameters. Their complexity allows them to learn different types of visual encodings in one training session. 
\end{itemize}

@@

\subsection{Results}

\begin{figure}[!t]
  \centering
    \includegraphics[width=\linewidth]{figure4_mlae_with_multi_and_humans_all.pdf}
  \caption{\textbf{Computational results of the position-length experiment.} \textit{Left:} Rasterized visualizations of type 1--5 for divided and grouped bar charts of Cleveland and McGill's position-length experiment. \textit{Right:} MLAE and $95\%$ confidence intervals for different regressors estimating the value of marked quantities in the visualizations. The VGG19 * and Xception * networks are using ImageNet weights. The top 5 rows represent networks trained on a single encoding while the last row shows `multi' networks which were trained on a random stream of types 1--5. We visually compare against human performance from the original experiment.}
  \label{fig:figure4_mlae}
\end{figure}

\noindent\textbf{Perceptual Performance.} In the original position-length experiment, Cleveland and McGill report that types 1-5 were post-ordered by their perceptual difficulty with type 1 being the easiest and type 5 the hardest. We report the average MLAE for each type across our networks (and total 56 runs per type) as follows: 
for type 1 $MLAE= 3.956 $ ($SD= 0.274 $),
for type 2 $MLAE= 3.952 $ ($SD= 0.441 $),
for type 3  $MLAE= 4.349 $ ($SD= 0.367 $),
for type 4 $MLAE= 3.668 $ ($SD= 0.256 $), and 
for type 5 $MLAE= 3.902 $ ($SD= 0.253 $). These distributions yield significance ($F_{4,25}=2.815, p<0.05$) but post hoc comparions show that really only type 3 and type 4 differ ($t_{10}=3.406, p<0.01$). This means that the networks do not prefer a certain type on average which leads us to \textbf{partially accept H2.1} and we do not replicate the same pattern as Cleveland and McGill in their human studies even though our networks' performance is clearly worse than their human baseline.
\\~\\
\noindent\textbf{`Multi' Network Performance.} In Cleveland and McGill's original position-length experiment, humans were asked to judge visualizations of types 1-5. In the bottom of Figure~\ref{fig:figure4_mlae} we average the human performances of types 1-5 to create a `multi' human. Similarly, when training our networks, we first limit each training to one stimuli. However, a `multi' network which learns types 1-5 simultaenously is closer to the human in the original user study.

>>>>

\section{Experiment 3: Position-Length}
\label{sec:positionlength}

Cleveland and McGill assess the perception of position and length across five designs of grouped and divided bar charts (Fig.~\ref{fig:figure4_mlae}). Both types of chart can show the same information, but the elementary perceptual task is different: a grouped bar chart always involves the judgment of positions along a common scale, while a divided bar chart also requires length judgments. Types 1, 2, and 3 involve relating the judgment of positions along a common scale while types 4 and 5 involve relating the measure of length. For each graph, two bars or bar segments were marked, and participants were asked to judge what percentage the smaller marked element was of the larger. Cleveland and McGill ordered the types from easiest (type 1) to hardest (type 5).

\setlength{\belowdisplayskip}{3pt} \setlength{\belowdisplayshortskip}{3pt}
\setlength{\abovedisplayskip}{3pt} \setlength{\abovedisplayshortskip}{3pt}

For data generation, we follow the same approach as in the original experiment. We generate ten value pairs using the following equation:
\begin{equation}
s_i = 10\times10^{(i-1)/12},\textnormal{~~~~~}i=1,...,10,
\end{equation}
Then, we generate eight other random values in the range of 10 and 93. These boundaries were chosen such that the largest first-layer convolutional filter size in our networks (of $5\times5$ in LeNet) would see all content in our $100\times100$ pixel image. The paired quantities/elements are marked by a single pixel. We ask our networks to estimate the ratio of the smaller to the larger, which we model as a single value regression problem. For type 4, we follow Cleveland and McGill's constraint that neither the top or the bottom of the marked quantities match, forcing estimations of length rather than position. This task has $9.20\times10^{16}$ possible permutations---a challenging problem for the capacity of our networks, but one that Cleveland and McGill found can be solved reliably by humans with $\approx6.5\%$ error \cite{cleveland_mcgill}.

%We model this experiment as a single value regression task and ask networks to estimate the percentage of the smaller to the larger marked quantity in each visual encoding. The network first has to find the two marked quantities, then identify the smaller one, and finally estimate the ratio in comparison to the larger quantity. The 8 random quantities (which should be ignored by the network) push the number of possible permutations to a massive $9.20E+16$ and creates a very challenging problem.

@@

\begin{hypolist}
  \item \textbf{H3.1} \textbf{Our networks can estimate all types equally well.} A grouped bar chart involves judging a position while a divided bar chart most likely (if not type 2) requires length judgments. Our rankings of elementary perceptual tasks do not yield a strong preference for either across all networks.
  \item \textbf{H3.2} \textbf{A trained multi-task network will work as well as individual trained networks.} We train a multi-task network (labeled `multi') from all five types. While we fix the number of trainable parameters to be the same as in the single task network, CNNs have a hierarchical structure which allows them to learn intermediate representations that are useful for multiple tasks.
\end{hypolist}

@@

\vspace{-0.1cm}
\subsection{Results}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.90\linewidth]{figure4_mlae_with_multi_and_humans_all_NEW.pdf}
  \caption{\textbf{Computational results of the position-length experiment.} \textit{Left:} Type 1--5 stimuli for divided and grouped bar charts (as per Cleveland and McGill). \textit{Right:} MLAE and \change{bootstrapped} $95\%$ confidence intervals of our networks. Star * denotes networks using ImageNet weights. The last row shows `multi' networks trained on a random stream of types 1--5 \change{(plus VGG19 with $5\times$ training data)}. \change{We include human performance (black) from the original experiment (ClMcG, top) and from Heer and Bostock's crowdsourced studies (HeerBos, bottom).}}
  \label{fig:figure4_mlae}
\end{figure}

%\JT{State random performance.} 
%\JT{Mention Heer and Bostock's results in body text.}

\noindent\textbf{Perceptual Performance.} \change{Midmean random performance is at $MLAE=4.7$.} Average MLAE across our networks is: 
type 1 $MLAE= 3.956 $ ($SD= 0.274 $),
type 2 $MLAE= 3.952 $ ($SD= 0.441 $),
type 3 $MLAE= 4.349 $ ($SD= 0.367 $),
type 4 $MLAE= 3.668 $ ($SD= 0.256 $), and 
type 5 $MLAE= 3.902 $ ($SD= 0.253 $). This yields significance ($F_{4,25}=2.815, p<0.05$), but post-hoc comparisons show that only types 3 and 4 differ ($t_{10}=3.406, p<0.01$). Thus, our networks do not prefer a certain type on average, and so we \textbf{partially accept H3.1}. %Further, we do not replicate the same type difficulty ordering as Cleveland and McGill in their human studies.

Overall, our networks' performances are clearly worse than their human baselines. The problem space in these visual relation tasks is much larger than in the elementary perceptual tasks, resulting in average errors of $12$--$20\%$. Further, the finding from the elementary perceptual tasks that position and length judgments had approximately equivalent rank is consistent with these findings.
\\~\\
\noindent\textbf{Multi-task Network Performance.} In the original position-length experiment, humans were asked to judge visualizations across types 1-5. In the last row of Fig.~\ref{fig:figure4_mlae}, we average the human performances to create an average score across tasks. For our multi-task networks trained on all stimuli types, we record an average error across all classifiers of $MLAE=4.358$ ($SD=0.327$). Then, we compare against the average errors for all types, as reported above for perceptual performance. We reach significant differences ($F_{5,30}=3.454,p<0.05$). Post-hoc comparisons yield significant differences between the multi-task network and type 4 ($t_{10}=3.716, p<0.01$) and also to type 5 ($t_{10}=2.467, p<0.05$). Since the average MLAE is worse than all of types 1--5, and the distributions observe significant differences, we acknowledge that the multi-type task is harder for the networks than learning single types of encodings. We test whether there is insufficient training data for this task by providing five times more training data; however, this decreases performance. Thus, we \textbf{reject H3.2}. One promising exception is VGG trained from scratch, though performance has a wide variance across cross-validation sets.

### 2018-08-15 PAPER/06_bars_and_framed_rectangles_experiment.tex
Update paper from overleaf.

\section{Experiment: Bars and Framed Rectangles}

Visual cues can help converting graphical elements back to their real world variables. Cleveland and McGill introduced the bars and framed rectangles experiment which judges the elementary perceptual task of position along non-aligned scales~\cite{cleveland_mcgill}.

\subsection{Hypotheses}

We proposed two hypotheses entering the elementary perceptual task experiment:

\begin{itemize}
	\item \textbf{H4.1} \textbf{Classifiers can leverage additional visual cues.} The original bar and framed rectangle experiment shows how visual cues aid humans in mapping graphical elements to quantitative variables. This should be the same for feed-forward neural networks since they are inspired by the visual system.
	\item \textbf{H4.2} \textbf{Weber's law can be transferred to computational perception.} Cleveland and McGill confirmed Weber's law based on the bar and framed rectangle experiment. For humans, the ability to perceive change within a distribution is proportional to the size of the initial distribution.
\end{itemize}

@@

@@

First run indicates that framed rectangles perform better but we dont really know it yet.

\begin{figure}[t]
	  \includegraphics[width=\linewidth]{figure12_mlae_better_all.pdf}
  \caption{\textbf{Computational results of the Bars-and-Framed-Rectangles experiment.} Log absolute error means and 95\% confidence intervals for the \emph{bars-and-framed-rectangles experiment} as described by Cleveland and McGill~\cite{cleveland_mcgill}. We test the performance of a Multi-layer Perceptron (MLP), the LeNet Convolutional Neural Network, as well as feature generation using the VGG19 and Xception networks trained on ImageNet.}
	\label{fig:figure12_mlae}
\end{figure}

>>>>

\section{Experiment 4: Bars and Framed Rectangles}
\label{sec:barsframedrectangles}

Visual cues can help in converting graphical elements back to their real-world variables. Cleveland and McGill introduced the bars-and-framed-rectangles experiment to compare the perceptual judgment of length and position along non-aligned scales. Fig.~\ref{fig:figure12_mlae} shows both variations on the left. Without framing, it is difficult to judge which bar is larger (bottom). However, with a frame showing maximum length, this length judgment \change{can be} converted into a position judgment along non-aligned scales, which simplifies the perceptual problem.

Cleveland and McGill theorize that judging the framed whitespace could be considered a length rather than a position judgment. Given this, they relate the task to Weber's Law: the perceivable difference within a distribution is proportional to the initial size of the distribution~\cite{householder1940weber}. For this experiment, Weber's Law implies that humans can more easily measure the difference in the white scale since its initial size is small, whereas estimating the small change in lengths of the black bars is harder. The Just Noticeable Difference (JND) is higher when the initial stimulus is smaller in size.

We set up the experiment as a two value regression task (Fig.~\ref{fig:figure12_mlae}). \change{Each bar length varies randomly by between 1 and 12 of 60 pixels, with each bar undergoing an individual vertical shift of up to 20 pixels. There are 132 different labels over 50,160 possible stimuli.}

%\subsection{Hypothesis}

\begin{hypolist}
	\item \textbf{H4.1} \textbf{Networks performance will improve with additional visual cues.} The original bar and framed rectangle setting shows how visual cues aid humans in mapping graphical elements to quantitative variables. This should be the same for feed-forward neural networks, as we give them more signal from which to learn.
\end{hypolist}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=0.9\linewidth]{figure12_mlae_better_all_NEW.pdf}
  \caption{\textbf{Computational results of the bars-and-framed-rectangles experiment.} \textit{Left:} Stimuli of two bars for length judgment (bottom) following Cleveland and McGill's setting. Perceiving which bar is longer is \change{significantly} easier for humans when a frame is added (top). \textit{Right:} For networks (trained from scratch, or * indicates ImageNet weights), there seems no significant difference between the encodings as reported by MLAE and \change{bootstrapped} 95\% confidence intervals.}
	\label{fig:figure12_mlae}
\end{figure}

@@

\change{Midmean random performance in this task is equal to $MLAE=4.8$.} We observe varying performance for our networks. Averaged across networks: the framed rectangle encoding $MLAE=1.982$ ($SD=0.89$) and for the bars encoding $1.867$ ($SD=0.709$). This difference was not significant, and so we \textbf{reject H4.1}. VGG19 again can regress the length in both cases, for the framed rectangle encoding $MLAE=0.595$ ($SD=0.225$) and for the bar encoding $MLAE=0.735$ ($SD=0.410$), though with higher variance without the added visual cues. The difference in relation to the visual cue is not significant. %\JT{Rejection of 4.1 calls into question the memorization of networks.}

\change{Further, our user study provides evidence that humans can more easily measure the frame rectangles. Participants were able to estimate the bar length with less error when frames are present $MLAE= 3.336 $ ($SD= 0.828 $) than without this visual aid $MLAE= 3.928 $ ($SD= 0.42 $). This difference is significant $F_{1,48}=9.765, p<0.01$.}

%\\~\\
%Unfortunately, both hypotheses were rejected for this experiment. However, it was great to see a partial validation of our ranking of elementary perceptual tasks (Table~\ref{tab:ranking}).
